---
title: "The Third Branch: Partisanship and Perception of the Supreme Court of the United States"
author: "Felicity Zhang"
date: "May 11, 2026"
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    colorlinks: true
    fig-pos: "H"
execute:
  echo: true
  warning: false
  message: false
---

<https://github.com/felicityxz17/gov51-finalproj>

# 1 Introduction

The judiciary has long been viewed as a vital pillar of American government, serving as the integral third branch within our delicate political system of checks and balances. At the head of the federal judiciary is the Supreme Court of the United States, responsible for delivering the final verdict on many of the nation's most contentious and consequential legal disputes around issues ranging from abortion to gun laws. With no elections or term limits and the power of judicial review, the Court holds the unique and vital role — at least in theory — of holding fast to unbiased legal interpretation, a bulwark against the fluctuating forces of politicization from the outside world.

However, even as the Supreme Court is designed to be shielded from the sway of public opinion, the trust and perception of American voters is simultaneously a critical mechanism to gauge how effectively the Court is performing its intended task. In recent years, public confidence in the Supreme Court has plummeted. The University of Pennsylvania's Annenberg Public Policy Center published a report in 2022 detailing that over half of the American public disapproves of the current Court, and the Pew Research Center found in 2025 that favorable views of the Court remain near these historic lows. These trends mirror both a broader decline in the perceived legitimacy of political institutions and a parallel growth in polarization and partisanship among American citizens. This paper examines the latter to answer the question: What, if any, is the association between voter partisanship and perceptions of the Supreme Court in today's political climate?

Studies on public perception of the Supreme Court are relatively rare and generally follow an overarching pattern that lacks particular emphasis on partisanship. Existing analysis, such as the aforementioned research by the Annenberg Public Policy Center and Pew Research Center, has identified that Democrats are more likely to disprove of the current Court, aligning with the current 6-3 partisan split of the Court between conservative and liberal justices. However, at present, the literature lacks a more thorough examination of the relationship between partisanship and perceptions of Court ideology and power. In particular, previous work focuses on longitudinal trends over detailed partisan breakdown, frequently using aggregation and simplification to present findings that only address the general Republican and Democrat blocs and thus make it difficult to discern whether there is a consistent directional trend between partisanship and perception, or whether voters at either extreme within each party are most responsible for the differences in perceived approval.

This paper visually investigates various aspects of voter perception of the Supreme Court, ranging from overall approval to ideology and power. Together, these descriptive analyses confirm a strong and consistent relationship between partisanship and approval of the Court. Voters who identify more strongly as Republican are more likely to rate the Court favorably in all aspects. However, Republicans approve of the Court less strongly than Democrats disapprove of it. Simultaneously, there is a strong bipartisan consensus against Supreme Court justices allowing personal political views to influence their decisions.

# 2 Data and Measurement

## 2.1 Dataset: ATP W149

This paper utilizes Wave 149 of the Pew Research Center's American Trends Panel survey, which was conducted from July 1 to July 7, 2024 via self-administered web survey or by life telephone interviewing and contains responses from a nationally-representative sample of 9,424 randomly-selected U.S. adults. Few publicly-available large-scale surveys contain multiple indicators of both Supreme Court perception and partisanship; therefore, although ATP W149 did not deal primarily with the Supreme Court, it contains a handful of relevant questions that are useful for this paper's analysis. Some of the data discussed here has been previously analyzed by the Pew Research Center; however, as previously discussed, this existing analysis focuses more on longitudinal trends rather than detailed partisan breakdown, and does not examine the implications of ATP W149's questions about how voters perceive the ideology and power of the Supreme Court that may provide more nuanced indications of the relationship between partisanship and Court perception.

At the time that ATP W149 was conducted, the Supreme Court consisted of 9 justices, with a conservative majority of 6 justices and a liberal minority of 3 justices. This context is important in subsequent data analysis as Court makeup is expected to significantly inform partisan perceptions, a dynamic that will be further discussed within this paper's limitations and conclusion sections.

## 2.2 Data Setup

```{r}
library(tidyverse)
library(knitr)
atp <- read_csv("data/raw/ATP W149.csv")
```

The ATP W149 dataset contains 9424 responses to 160 total questions, of which only a handful are relevant to this project. Because of the large dataset, a preview of the data has been omitted for simplicity. The relevant questions are filtered below. The full survey questionnaire is available for reference in the \`data\` folder.

```{r}
atp <- atp %>%
  mutate(party = case_when(
    F_PARTY_FINAL == 1 ~ 1, 
    F_PARTY_FINAL == 2 ~ 5,
    (F_PARTY_FINAL == 3 | F_PARTY_FINAL == 4 | F_PARTY_FINAL == 99) ~ case_when(F_PARTYLN_FINAL == 1 ~ 2, F_PARTYLN_FINAL == 2 ~ 4, F_PARTYLN_FINAL == 99 ~ 3)))
```

Two questions within the dataset directly elicit the partisan affiliation of respondents. Respondents are first asked which political party they identify with, and those who respond with "Independent," "Something else," or refuse to answer are then asked a second question of which political party they lean more towards. For the purposes of this paper, responses to these two questions are reassigned to a 5-point scale, where respondents who choose Republican/Democrat on the first question are classified as 1/5, respectively; respondents who choose Lean Republican/Lean Democrat on the second question are classified as 2/4, respectively; and respondents who respond with "Independent," "Something else," or refuse to answer the first question AND refuse to answer the second question are classified as 3. The creation of this 5-point scale thus enables a more detailed breakdown that simplifies analysis across political parties.

```{r}
byparty <- data.frame(summarize(group_by(atp, party), n()))

knitr::kable(byparty, col.names = c("Party Code", "Number of Respondents"), caption = "Respondent Party Breakdown")
```

The modified dataset contains 2666 "Republican", 1702 "Lean Republican", 255 "Moderate", 1851 "Lean Democrat", and 2950 "Democrat" respondents. The number of (Lean) Republicans and (Lean) Democrats are relatively symmetrical, suggesting that the dataset is relatively balanced.

# 3 Overall Approval

```{r}
court_opinion <- atp %>%
  filter((INSTFAV_COURT_W149 != 98) & (INSTFAV_COURT_W149 != 99))

length(court_opinion$INSTFAV_COURT_W149)

court_opinion_byparty <- data.frame(summarize(group_by(court_opinion, party), n()))

knitr::kable(court_opinion_byparty, col.names = c("Party Code", "Number of Respondents"), caption = "Respondent Party Breakdown for Overall Approval of the Supreme Court")
```

The first aspect of analysis is the standard overall approval of the Supreme Court. ATP W149 asks respondents a multi-part question, "Do you have a favorable or unfavorable opinion on each of the following?," with the Supreme Court sub-question as the question of interest for this paper. Respondents are asked to provide a rating from 1 to 4, with 1 representing "Very favorable," 2 representing "Mostly favorable, 3 representing "Mostly unfavorable, and 4 representing "Very unfavorable." For the purposes of this analysis, non-responses (coded as 98 or 99 in the data) are excluded, leaving the above distribution of partisan affiliation among 9308 respondents that roughly resembles that of the original dataset.

```{r}
court_opinion_party1 <- court_opinion %>% filter(party == 1)
court_opinion_party2 <- court_opinion %>% filter(party == 2)
court_opinion_party3 <- court_opinion %>% filter(party == 3)
court_opinion_party4 <- court_opinion %>% filter(party == 4)
court_opinion_party5 <- court_opinion %>% filter(party == 5)

court_opinion_1 <- c(
  mean(court_opinion_party1$INSTFAV_COURT_W149 == 1),
  mean(court_opinion_party2$INSTFAV_COURT_W149 == 1),
  mean(court_opinion_party3$INSTFAV_COURT_W149 == 1),
  mean(court_opinion_party4$INSTFAV_COURT_W149 == 1),
  mean(court_opinion_party5$INSTFAV_COURT_W149 == 1)
)
court_opinion_2 <- c(
  mean(court_opinion_party1$INSTFAV_COURT_W149 == 2),
  mean(court_opinion_party2$INSTFAV_COURT_W149 == 2),
  mean(court_opinion_party3$INSTFAV_COURT_W149 == 2),
  mean(court_opinion_party4$INSTFAV_COURT_W149 == 2),
  mean(court_opinion_party5$INSTFAV_COURT_W149 == 2)
)
court_opinion_3 <- c(
  mean(court_opinion_party1$INSTFAV_COURT_W149 == 3),
  mean(court_opinion_party2$INSTFAV_COURT_W149 == 3),
  mean(court_opinion_party3$INSTFAV_COURT_W149 == 3),
  mean(court_opinion_party4$INSTFAV_COURT_W149 == 3),
  mean(court_opinion_party5$INSTFAV_COURT_W149 == 3)
)
court_opinion_4 <- c(
  mean(court_opinion_party1$INSTFAV_COURT_W149 == 4),
  mean(court_opinion_party2$INSTFAV_COURT_W149 == 4),
  mean(court_opinion_party3$INSTFAV_COURT_W149 == 4),
  mean(court_opinion_party4$INSTFAV_COURT_W149 == 4),
  mean(court_opinion_party5$INSTFAV_COURT_W149 == 4)
)

court_opinion_comparison <- cbind(court_opinion_1, court_opinion_2, court_opinion_3, court_opinion_4)
barplot(court_opinion_comparison, beside = T,
        col = c("red", "pink", "white", "skyblue", "blue"),
        names.arg = c("1 (Very favorable)", "2 (Mostly favorable)", "3 (Mostly unfavorable)", "4 (Very unfavorable)"),
        cex.names = 0.6,
        ylab = "Party Proportion",
        main = "Figure 1: Supreme Court Overall Approval by Party"
          )

court_opinion_proportions <- data.frame(c("1 (Republican)", "2 (Lean Republican)", "3 (Moderate)", "4 (Lean Democrat)", "5 (Democrat)"), court_opinion_comparison)
knitr::kable(court_opinion_proportions, col.names = c("Party", "1 (Very favorable)", "2 (Mostly favorable)", "3 (Mostly unfavorable)", "4 (Very unfavorable)"), caption = "Proportion of Each Political Party by Approval Rating of the Supreme Court")
```

Figure 1 is a side-by-side bar chart comparing the proportions of respondents providing each rating within their party. There is a clear relationship between partisan affiliation and approval, with respondents who identify more toward the Democrat end of the spectrum displaying lower favorability ratings, while respondents who identify more toward the Republican end of the spectrum display higher favorability ratings. However, consistent with the overall low approval of the Court, Republicans appear less strongly approving of the Court than Democrats appear approving, as the majority (62% and 56%) of Republicans and Lean Republicans rate the Court "Mostly favorable," while around 50% of Democrats and Lean Democrats rate the Court as not merely "Mostly unfavorable" but "Very unfavorable."

```{r}
court_opinion_summary <- data.frame(summarize(group_by(court_opinion, party), mean(INSTFAV_COURT_W149), sd(INSTFAV_COURT_W149)))

knitr::kable(court_opinion_summary, col.names = c("Party", "Mean", "Standard Deviation"), caption = "Summary Statistics of Supreme Court Approval Overall by Party")

lm_court_opinion = lm(INSTFAV_COURT_W149 ~ party, data = court_opinion)
summary(lm_court_opinion)
```

Both the summary statistics table and linear model output confirm a clear positive association between `party` value on the 5-point scale and unfavorability rating of the Court, i.e., increasing Democrat leaning correlates with increasing perceived unfavorability. Although the interpretation of the linear model is not entirely accurate because both variables are technically categorical rather than numerical, the extremely low t-statistic suggests that there is some statistically significant association between party and favorability. Furthermore, the R-squared of the model is 0.3404, suggesting that party affiliation alone (roughly, when measured on a scale of 1-5) explains a noticeable 34% of variability in favorability ratings.

# 4 Perceived Power

```{r}
court_power <- atp %>%
  filter((COURTPOWER_W149 != 98) & (COURTPOWER_W149 != 99))

# Reordering
court_power <- court_power %>%
  mutate(COURTPOWER_W149_MOD = case_when(
    COURTPOWER_W149 == 1 ~ 3,
    COURTPOWER_W149 == 2 ~ 1,
    COURTPOWER_W149 == 3 ~ 2
  ))

length(court_power$COURTPOWER_W149_MOD)

court_power_byparty <- data.frame(summarize(group_by(court_power, party), n()))

knitr::kable(court_power_byparty, col.names = c("Party Code", "Number of Respondents"), caption = "Respondent Party Breakdown for Perceived Power of the Supreme Court")
```

Although the perceived power of the Supreme Court is a less conventional measure of Supreme Court approval, it offers an alternative angle that provides valuable insight into voters' perceptions of the Supreme Court. ATP W149 asks respondents, "Do you think the U.S. Supreme Court has...," with responses of 1, 2, 3, and 98/99 indicating "Too much power," "Too little power," "Right amount of power," and N/A/unanswered, respectively. For clearer comparison, these responses were reordered such that the numerical values of 1, 2, and 3 represent "Too little power," "Right amount of power," and "Too much power," respectively. Again, non-responses are excluded, leaving 4601 responses that roughly retain the original cross-party distribution.

```{r}
court_power_party1 <- court_power %>% filter(party == 1)
court_power_party2 <- court_power %>% filter(party == 2)
court_power_party3 <- court_power %>% filter(party == 3)
court_power_party4 <- court_power %>% filter(party == 4)
court_power_party5 <- court_power %>% filter(party == 5)

court_power_1 <- c(
  mean(court_power_party1$COURTPOWER_W149_MOD == 1),
  mean(court_power_party2$COURTPOWER_W149_MOD == 1),
  mean(court_power_party3$COURTPOWER_W149_MOD == 1),
  mean(court_power_party4$COURTPOWER_W149_MOD == 1),
  mean(court_power_party5$COURTPOWER_W149_MOD == 1)
)
court_power_2 <- c(
  mean(court_power_party1$COURTPOWER_W149_MOD == 2),
  mean(court_power_party2$COURTPOWER_W149_MOD == 2),
  mean(court_power_party3$COURTPOWER_W149_MOD == 2),
  mean(court_power_party4$COURTPOWER_W149_MOD == 2),
  mean(court_power_party5$COURTPOWER_W149_MOD == 2)
)
court_power_3 <- c(
  mean(court_power_party1$COURTPOWER_W149_MOD == 3),
  mean(court_power_party2$COURTPOWER_W149_MOD == 3),
  mean(court_power_party3$COURTPOWER_W149_MOD == 3),
  mean(court_power_party4$COURTPOWER_W149_MOD == 3),
  mean(court_power_party5$COURTPOWER_W149_MOD == 3)
)

court_power_comparison <- cbind(court_power_1, court_power_2, court_power_3)
barplot(court_power_comparison, beside = T,
        col = c("red", "pink", "white", "skyblue", "blue"),
        names.arg = c("1 (Too little power)", "2 (Right amount of power)", "3 (Too much power)"),
        cex.names = 0.7,
        ylab = "Party Proportion",
        main = "Figure 2: Supreme Court Perceived Power by Party"
          )

court_power_summary <- data.frame(summarize(group_by(court_power, party), mean(COURTPOWER_W149_MOD), sd(COURTPOWER_W149_MOD)))

court_power_proportions <- data.frame(c("1 (Republican)", "2 (Lean Republican)", "3 (Moderate)", "4 (Lean Democrat)", "5 (Democrat)"), court_power_comparison)
knitr::kable(court_power_proportions, col.names = c("Party", "1 (Too little power)", "2 (Right amount of power)", "3 (Too much power)"), caption = "Proportion of Each Political Party by Perceived Power of the Supreme Court")


knitr::kable(court_power_summary, col.names = c("Party", "Mean", "Standard Deviation"), caption = "Summary Statistics of Supreme Court Perceived Power by Party")
```

Because a linear model would be even less interpretable in this case, where the three limited categories of perceived power do not offer room for the sliding-scale interpretation of favorability, we address only the side-by-side barchart in Figure 2 and the summary statistics table. There is once again a clear, consistent trend where the more Republican a voter leans, the more likely they are to perceive the Court as having the right amount of power compared to Democrats; similarly, the more Democrat a voter leans, the more likely to perceive the Court as having too much power. Notably, a very small minority across all parties perceive the Court as having too little power; the debate is centered entirely around whether the Court has an appropriate or excessive amount of power. Also, in contrast to Figure 1, Democrats and Lean Democrats appear slightly less likely to rate the Court as having too much power (71% and 64%) than Republicans and Lean Republicans appear likely to rate the Court as having the right amount of power (79% and 71%), and Democrats appear slightly more likely to rate the Court as having the right amount of power (31 and 26%) than Republicans appear likely to rate the Court as having too much power (17 and 24%). This discrepancy might be reasonably interpreted as evidence that voters who do not align with the current partisan makeup of the Court disapprove of the Court's decisions and actions because of an inconsistency with their personal beliefs, but that they simultaneously understand that this inconsistency is inevitable (and temporary) and therefore should not reflect poorly on the Court as an institution. Therefore, the overall approval rating of the Supreme Court might be interpreted as more susceptible to partisan beliefs because it addresses general perceptions toward the Court that might be informed by narrower opinions about individual justices or decisions. Consequently, low public approval should not necessarily be interpreted as evidence to support the belief that the Supreme Court as an institution is failing and in need of reform.

# 5 Court Politicization

A third and final angle to measure perception and partisanship of the Supreme Court relates to voters' perceptions of how partisan the Supreme Court itself ought to be, and how well the Supreme Court is matching this intended level of partisanship. ATP W149 asks respondents both of these questions, which will be examined in turn.

## 5.1 Prescriptive Court Politicization

```{r}
court_politpresc <- atp %>%
  filter((SCOTUS_PRTSN_W149 != 98) & (SCOTUS_PRTSN_W149 != 99))

length(court_politpresc$SCOTUS_PRTSN_W149)

court_politpresc_byparty <- data.frame(summarize(group_by(court_politpresc, party), n()))

knitr::kable(court_politpresc_byparty, col.names = c("Party Code", "Number of Respondents"), caption = "Respondent Party Breakdown for Prescriptive Politicization in the Supreme Court")
```

ATP W149 first asks respondents the prescriptive question, "Thinking about the Supreme Court, do you think justices should or should not bring their own political views into how they decide major cases?," with responses of 1, 2, and 98/99 indicating "Should do this," "Should NOT do this," and N/A/unanswered, respectively. Once again, non-responses are excluded, leaving 9287 responses that roughly retain the original cross-party distribution.

```{r}
court_politpresc_party1 <- court_politpresc %>% filter(party == 1)
court_politpresc_party2 <- court_politpresc %>% filter(party == 2)
court_politpresc_party3 <- court_politpresc %>% filter(party == 3)
court_politpresc_party4 <- court_politpresc %>% filter(party == 4)
court_politpresc_party5 <- court_politpresc %>% filter(party == 5)

court_politpresc_1 <- c(
  mean(court_politpresc_party1$SCOTUS_PRTSN_W149 == 1),
  mean(court_politpresc_party2$SCOTUS_PRTSN_W149 == 1),
  mean(court_politpresc_party3$SCOTUS_PRTSN_W149 == 1),
  mean(court_politpresc_party4$SCOTUS_PRTSN_W149 == 1),
  mean(court_politpresc_party5$SCOTUS_PRTSN_W149 == 1)
)
court_politpresc_2 <- c(
  mean(court_politpresc_party1$SCOTUS_PRTSN_W149 == 2),
  mean(court_politpresc_party2$SCOTUS_PRTSN_W149 == 2),
  mean(court_politpresc_party3$SCOTUS_PRTSN_W149 == 2),
  mean(court_politpresc_party4$SCOTUS_PRTSN_W149 == 2),
  mean(court_politpresc_party5$SCOTUS_PRTSN_W149 == 2)
)

court_politpresc_comparison <- cbind(court_politpresc_1, court_politpresc_2)
barplot(court_politpresc_comparison, beside = T,
        col = c("red", "pink", "white", "skyblue", "blue"),
        names.arg = c("1 (Should)", "2 (Should NOT)"),
        cex.names = 0.8,
        ylab = "Party Proportion",
        main = "Figure 3: Should Supreme Court justices' personal political views \n affect their decisions?"
          )

court_politpresc_summary <- data.frame(summarize(group_by(court_politpresc, party), mean(SCOTUS_PRTSN_W149), sd(SCOTUS_PRTSN_W149)))

court_politpresc_proportions <- data.frame(c("1 (Republican)", "2 (Lean Republican)", "3 (Moderate)", "4 (Lean Democrat)", "5 (Democrat)"), court_politpresc_comparison)
knitr::kable(court_politpresc_proportions, col.names = c("Party", "1 (Should)", "2 (Should NOT)"), caption = "Proportion of Each Political Party by Whether the Supreme Court Should Be Affected by Personal Political Views By Party")

knitr::kable(court_politpresc_summary, col.names = c("Party", "Mean", "Standard Deviation"), caption = "Summary Statistics of Whether the Supreme Court Should Be Affected by Personal Political Views By Party")
```

Figure 3 displays a notable difference from the previous two barcharts: it is consistent regardless of party. Thus, there is a broad bipartisan consensus — indeed, around 90% across all parties — that Supreme Court justices' personal political view should NOT affect their decisions. The barchart and summary statistics table still display a slight consistent correlation between partisanship and perception, with increasing Democrat leaning correlating with a higher likelihood of believing that justices' personal political views should *not* affect their decisions. An ANOVA F-test might determine whether this difference is statistically significant but will not be conducted here.

## 5.2 Descriptive Court Politicization

```{r}
court_politdesc <- atp %>%
  filter(SCOTUS_JOB_W149 != 99)

# Reordering
court_politdesc <- court_politdesc %>%
  mutate(SCOTUS_JOB_W149_MOD = case_when(
    SCOTUS_JOB_W149 == 1 ~ 1,
    SCOTUS_JOB_W149 == 2 ~ 2,
    SCOTUS_JOB_W149 == 3 ~ 4,
    SCOTUS_JOB_W149 == 4 ~ 5,
    SCOTUS_JOB_W149 == 5 ~ 3
  ))

length(court_politdesc$SCOTUS_JOB_W149_MOD)

court_politdesc_byparty <- data.frame(summarize(group_by(court_politdesc, party), n()))

knitr::kable(court_politdesc_byparty, col.names = c("Party Code", "Number of Respondents"), caption = "Respondent Party Breakdown for Descriptive Politicization in the Supreme Court")
```

Based on their responses to the previous prescriptive question, ATP W149 asks (the majority of) respondents who responded that justice should NOT bring their personal political views into their decisions the descriptive question, "How would you rate the job Supreme Court justices are doing in keeping their own political views out of how they decide major cases?," with responses of 1, 2, 3, 4, 5, and 98/99 indicating "Excellent," "Good," "Only fair," "Poor," "Not sure," and N/A/unanswered, respectively. Again, for clearer comparison, these responses were reordered such that the numerical values of 1, 2, 3, 4, and 5 represent "Excellent," "Good," "Not sure," "Only fair," and "Poor," respectively. Once again, non-responses are excluded, leaving 8469 responses that roughly retain the original cross-party distribution.

```{r}
court_politdesc_party1 <- court_politdesc %>% filter(party == 1)
court_politdesc_party2 <- court_politdesc %>% filter(party == 2)
court_politdesc_party3 <- court_politdesc %>% filter(party == 3)
court_politdesc_party4 <- court_politdesc %>% filter(party == 4)
court_politdesc_party5 <- court_politdesc %>% filter(party == 5)

court_politdesc_1 <- c(
  mean(court_politdesc_party1$SCOTUS_JOB_W149_MOD == 1),
  mean(court_politdesc_party2$SCOTUS_JOB_W149_MOD == 1),
  mean(court_politdesc_party3$SCOTUS_JOB_W149_MOD == 1),
  mean(court_politdesc_party4$SCOTUS_JOB_W149_MOD == 1),
  mean(court_politdesc_party5$SCOTUS_JOB_W149_MOD == 1)
)
court_politdesc_2 <- c(
  mean(court_politdesc_party1$SCOTUS_JOB_W149_MOD == 2),
  mean(court_politdesc_party2$SCOTUS_JOB_W149_MOD == 2),
  mean(court_politdesc_party3$SCOTUS_JOB_W149_MOD == 2),
  mean(court_politdesc_party4$SCOTUS_JOB_W149_MOD == 2),
  mean(court_politdesc_party5$SCOTUS_JOB_W149_MOD == 2)
)
court_politdesc_3 <- c(
  mean(court_politdesc_party1$SCOTUS_JOB_W149_MOD == 3),
  mean(court_politdesc_party2$SCOTUS_JOB_W149_MOD == 3),
  mean(court_politdesc_party3$SCOTUS_JOB_W149_MOD == 3),
  mean(court_politdesc_party4$SCOTUS_JOB_W149_MOD == 3),
  mean(court_politdesc_party5$SCOTUS_JOB_W149_MOD == 3)
)
court_politdesc_4 <- c(
  mean(court_politdesc_party1$SCOTUS_JOB_W149_MOD == 4),
  mean(court_politdesc_party2$SCOTUS_JOB_W149_MOD == 4),
  mean(court_politdesc_party3$SCOTUS_JOB_W149_MOD == 4),
  mean(court_politdesc_party4$SCOTUS_JOB_W149_MOD == 4),
  mean(court_politdesc_party5$SCOTUS_JOB_W149_MOD == 4)
)
court_politdesc_5 <- c(
  mean(court_politdesc_party1$SCOTUS_JOB_W149_MOD == 5),
  mean(court_politdesc_party2$SCOTUS_JOB_W149_MOD == 5),
  mean(court_politdesc_party3$SCOTUS_JOB_W149_MOD == 5),
  mean(court_politdesc_party4$SCOTUS_JOB_W149_MOD == 5),
  mean(court_politdesc_party5$SCOTUS_JOB_W149_MOD == 5)
)

court_politdesc_comparison <- cbind(court_politdesc_1, court_politdesc_2, court_politdesc_3, court_politdesc_4, court_politdesc_5)
barplot(court_politdesc_comparison, beside = T,
        col = c("red", "pink", "white", "skyblue", "blue"),
        names.arg = c("1 (Excellent)", "2 (Good)", "3 (Not sure)", "4 (Only fair)", "5 (Poor)"), 
        cex.names = 0.7,
        ylab = "Party Proportion",
        main = "Figure 4: How well are Supreme Court justices keeping \n their personal political views from \n affecting their decisions?"
          )

court_politdesc_proportions <- data.frame(c("1 (Republican)", "2 (Lean Republican)", "3 (Moderate)", "4 (Lean Democrat)", "5 (Democrat)"), court_politdesc_comparison)
knitr::kable(court_politdesc_proportions, col.names = c("Party", "1 (Excellent)", "2 (Good)", "3 (Not sure)", "4 (Only fair)", "5 (Poor)"), caption = "Proportion of Each Political Party by Whether the Supreme Court Is Affected by Personal Political Views By Party")

court_politdesc_summary <- data.frame(summarize(group_by(court_politdesc, party), mean(SCOTUS_JOB_W149_MOD), sd(SCOTUS_JOB_W149_MOD)))

knitr::kable(court_politdesc_summary, col.names = c("Party", "Mean", "Standard Deviation"), caption = "Summary Statistics of Whether the Supreme Court Is Affected by Personal Political Views By Party")
```

From Figure 4 and the summary statistics table, it is evident that despite bipartisan consensus on whether justices ought to be nonpartisan, voters' perceptions of whether judges are actually partisan reverts to the highly politicized trends from before. Leaning Republican is consistently correlated with a lower numerical response representing a more positive perception of justices' nonpartisanship. However, as in Figure 1, Figure 4 provides even more striking evidence that Democrats disapprove more strongly of the Court than Republicans approve. The majority (72% and 66%) of Democrats and Lean Democrats choose the lowest option, rating justices as doing a poor job of being nonpartisan. Meanwhile, Republicans are much more divided and hesitant to mirror such high approval, with a small minority (7% and 5%) of Republicans and Lean Republicans choosing the opposite rating of "Excellent," 40% and 26% choosing "Good," and a substantial proportion (29% and 35%) choosing "Only fair." The proportion of Republicans who agree with Democrats that justices are doing a poor job of being nonpartisan is around 11% and 18%, nearly as much as the proportion of Republicans who are not sure and greater than the proportion who think the justices are doing an excellent job. Thus, although Republicans do not condemn the Court's partisanship as strongly and severely as Democrats do, they appear somewhat hesitant and similarly skeptical that the Court is effectively remaining as nonpartisan as they believe it ought to be.

```{r}
lm_court_politdesc = lm(SCOTUS_JOB_W149_MOD ~ party, data = court_politdesc)
summary(lm_court_politdesc)
```

As in Section 3, although the linear model cannot be cleanly interpreted because it would require extrapolating categorical variables to continuous values, the model's positive and statistically significant slope coefficient supports the visual findings of the summary statistics table and barplots in confirming a clear positive association between `party` value on the 5-point scale and approval of nonpartisanship, i.e., increasing Democrat leaning correlates with increasing perceived (negative) partisanship. As before, the R-squared of the model is 0.3125, suggesting that party affiliation alone (roughly, when measured on a scale of 1-5) explains a noticeable 31% of variability in favorability ratings.

# Limitations

There are a number of limitations to the analysis conducted above, primarily owing to the nature of the original dataset. Firstly, due to the nature of the survey questions as providing limited categorical options rather than continuous numerical variables, there were limited statistics that could be derived and interpreted, necessitating primary reliance on the visual barcharts, and the statistics that were used such as means of ratings and coefficients of linear models were not always fully valid in their interpretation. However, they still provide meaningful indicators of the general directional trend of relationships between partisanship and perceptions, supported visually by the comparative barcharts.

A second area in which interpretive liberties may have been taken was with the (re)assignment of certain variables. Specifically, in the initial assignment of the 1-5 partisanship scale, respondents who refused to answer both questions were classified as being moderate, which may not be entirely accurate given that some respondents may have refused for other reasons, such as being unwilling to disclose their affiliation despite feeling strongly partisan or accidentally missing/skipping the questions. Similarly, respondents who initially answered that they considered themselves a Republican or a Democrat may not have been aware that there would have been a second question to specify their leanings had they omitted to answer the first, meaning that individuals who might otherwise be more moderate within their party might be slightly miscategorized between the "Republican/Democrat" and "Lean Republican/Democrat" categories. Additionally, the interpretation in Figure 4 of "Not sure" may not necessarily have been interpreted by respondents as representing a middle option between "Good" and "Only fair," although it was categorized as such within the visual analysis. Despite these limitations, the resulting data appears to visually support the choices made, as all relationships appear to be directionally consistent based on the chosen categories and orderings (e.g., the proportions of Democrat/Lean Democrat rating "Not sure" in Figure 4 is precisely in between the respective proportions rating "Good" and "Only fair"). These limitations also reflect a general, inevitable limitation of attempting to measure partisanship and public perception as variables that are neither continuous nor easily quantifiable. Thus, as with the analysis conducted here, the nature of the data may necessitate more general, abstract interpretations that emphasize the direction of trends rather than more precise, numerical ones.

Third, the sample sizes fluctuated slightly across questions due to the elimination of non-responses and N/As, which might affect the validity or robustness of certain results. Specifically, the question about perceived power of the Court has roughly half of the sample size of the other three questions. However, given that the ATP W149 survey is designed overall to be a representative sample of the U.S. population, that there do not appear to be any clear reasons for selection bias (e.g., it is unclear whether or why a voter of a particular partisan orientation would be motivated to skip the perceived power question), and that the sample sizes are still quite large (still over 4,000 for the perceived power question), this limitation should not be expected to significantly decrease the validity of the findings.

Finally, and perhaps most importantly, this study is purely descriptive and does not address or suggest any explanations for why the relationship between partisanship and perception is so strong. An obvious and intuitive explanation relates to the current makeup of the Court as a 6-3 conservative majority, as it aligns with the higher levels of Republican approval found consistently throughout this study. However, without any evidence to determine a causal relationship, this explanation must be offered merely as a possibility.

Future research might attempt to address these limitations by creating finer answer options, such as rating one's partisan orientation or one's perception of the Court's power on a sliding scale from 1-10. Similarly, future research might attempt to confirm whether there is indeed a relationship between partisan approval and the partisan split of the Court using temporal analyses and more long-term data comparing Court make-up to approval by party. However, ATP W149 is one of the only, if not the only, of recent surveys that has meaningfully asked about both partisan affiliation and these various aspects of Court approval. Therefore, future research might require more detailed public surveys to be conducted to have sufficient evidence for stronger conclusions.

# Conclusion

Based on a descriptive analysis of voters' perceptions of the Supreme Court, this paper finds that the current voting population is more likely to approve of the current Court if they identify as Republican and more likely to disapprove if they identify as Democrat, and that this relationship appears to operate analogous to a sliding scale with consistent directional change as one moves from the Republican to the Democrat end of the political spectrum. However, the strong bipartisan consensus supporting that the Court ought to be a nonpartisan institution, combined with the weaker approval of Republicans compared to the strong disapproval of Democrats, suggests that the current low approval of the Court reflects a general consensus that the Court is not acting as nonpartisan as it should, leading to dissatisfaction that is further exacerbated by partisan differences. Understanding the source of public disapproval of the Supreme Court is crucial to understanding whether, and how, it ought to be reformed.

# References

Copeland, Joseph. “Favorable Views of Supreme Court Remain near Historic Low.” *Pew Research Center*, 3 Sept. 2025, www.pewresearch.org/short-reads/2025/09/03/favorable-views-of-supreme-court-remain-near-historic-low/.

Maroni, Emily. “Over Half of Americans Disapprove of Supreme Court as Trust Plummets.” *The Annenberg Public Policy Center of the University of Pennsylvania*, 10 Oct. 2022, www.annenbergpublicpolicycenter.org/over-half-of-americans-disapprove-of-supreme-court-as-trust-plummets/.

Data from Pew Research Center American Trends Panel Wave 149, <https://www.pewresearch.org/dataset/american-trends-panel-wave-149/>.
